---
title: "NNC Thresholds"
author: "Gabriel Herrick"
date: "March 21, 2015"
output: html_document
---

If we consider all lakes, and look for the highest allowable values for each in the state of FL, we find that the maximum N and P values for a lake are Nmax = 2.23 mg/L and Pmax = 0.49 mg/L. Thus, the NNC Nmax:Pmax ratio is 2.23:0.49 = 4.55:1.  We have a very similar, constant ratio Ntreat:Ptreat = 2.240:0.496 = 4.52. Because our N:P ratio is constant in our treatments and commensurate with the NNC N:P ratio, we can use either N or P to calculate the nutrient loading in our treatments as a proportion of the maximum allowable nutrients according to the Numeric Nutrient Criteria (NNCmax). Thus, our High treatment = 2.24/2.23 = 1.00 NNCmax; our Low treatment = 0.42/2.23 = 0.188 NNCmax. Our control of plain tap water will be considered 0 mg/L = 0 NNCmax. 

# Nutrient Levels

```{r}
library(ggplot2)

treat.high.N <- 2.24
treat.low.N <- 0.42
treat.high.P <- 0.50
treat.low.P <- 0.093

NNC.N <- 2.23
NNC.P <- 0.49


dat <- data.frame(
  treatment= factor(c("Tap", "Low", "High"), levels = c("Tap", "Low", "High")),
  Nitrogen = c(0, treat.low.N, treat.high.N), 
  Phosphorus = c(0, treat.low.P, treat.high.P)
  )


```

Bar Graphs
```{r}
# very basic bar graph
ggplot(data = dat, aes(x=treatment, y=Nitrogen))+
  geom_bar(stat="identity")

# map treatment to different fill colors, add black line, no legend
ggplot(data = dat, aes(x=treatment, y=Nitrogen, fill = treatment))+
  geom_bar(color="black", stat="identity")+
  guides(fill=FALSE)

```
To draw multiple lines, the points must be grouped by a variable; otherwise all points will be connected by a single line. In this case, we want them to be grouped by nutrient. This means we need to modify the data frame.

```{r}
Nitrogen <- c(0, treat.low.N, treat.high.N)
Phosphorus <- c(0, treat.low.P, treat.high.P)
N.rel <- (Nitrogen/NNC.N)*100
P.rel <- (Phosphorus/NNC.P)*100

data <- data.frame(
  treatment= factor(rep(c("Tap", "Low", "High"),2), levels = c("Tap", "Low", "High")),
  nutrient = factor(rep(c("nitrogen", "phosphorus"), each=3)),
  concentration = c(N.rel,P.rel)
  )

```

Now that the data frame is modified, we can make the line graph with multiple lines. 

```{r}
# Basic line graph

line.plot <- ggplot(data=data, aes(x=treatment, y=concentration, group=nutrient)) +
    geom_line(aes(color=nutrient), size=2) +
    geom_point(size=3, fill="white")+
    xlab("Nutrient level") + ylab("Relative concentration")+
    geom_hline(aes(yintercept=100), size = 2)+
    #theme(axis.title.x=element_text(size=20))+
  #theme(axis.text.x = element_text(size=20))+
    #theme(axis.title.y=element_text(size=20))+
  #theme(axis.text.y = element_text(size=20))+
  annotate("text", x = 1.5, y = 95, label = "NNC threshold")+
  theme_bw(base_size=20)+
  theme(legend.position=c(.85, .50))
  
line.plot

  
````

Print line plot to file
```{r}
png("TreatmentLevels.png")
# or tiff("plot.tiff")
print(line.plot)
dev.off()

```



